Simple repression draft

### 2020-10-16 doc/resubmission/appendix_01_nonbursty.tex
fix ?? issues

Having derived an example chemical master equation we now focus on writing a
general matrix form for the kinetic models 1-4 shown in
Figure~\ref{fig1:means_cartoons}(C) in the main text. In each of these four
models, the natural mathematicization of their cartoons is as a chemical master
equation such as Eq.~\ref{eq:2state_rep_cme} for model 1, which for
completeness we reproduce here as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\label{eq:poisson_promoter_cme_appdx}
\end{split}
\end{align}
Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding the system with
$m$ mRNA molecules at time $t$ either in the repressor bound or unbound states,
respectively. $r$ is the probability per unit time that a transcript will be
initiated when the repressor is unbound, and $\gamma$ is the probability per
unit time for a given mRNA to be degraded. $k_R^-$ is the probability per unit
time that a bound repressor will unbind, while $k_R^+$ is the probability per
unit time that an unbound operator will become bound by a repressor. Assuming
mass action kinetics, $k_R^+$ is proportional to repressor copy number $R$.

@@

Although we derived Eq.~\ref{eq:generic_cme_appdx} for the particular case of
kinetic model 2 in Figure~\ref{fig1:means_cartoons}, in fact the
chemical master equations for all of the kinetic models in
Figure~\ref{fig1:means_cartoons} except for model 5 can be cast in this form.
(We treat model 5 separately in Appendix~\ref{sec:gen_fcn_appdx}.) Model
3 introduces no new subtleties beyond model 2 and
Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply with different
matrices of dimension four instead of three. Models 1 and 4 are both
handled by Eq.~\ref{eq:2state_rep_cme} in the main text, which is just
Eq.~\ref{eq:generic_cme_appdx} except in the special case of $\mathbf{R_D} =
\mathbf{R_A} \equiv \mathbf{R}$, since in these two models transcription
initiation events do not change promoter state.

@@

\subsubsection{Fano factor}
To compute the Fano factor, we first remove the repressor bound state from the
matrices describing the model, which reduce to
\begin{equation}
\mathbf{K} = \begin{pmatrix}
                -k_P^+ & k_P^- \\
                 k_P^+ &-k_P^- 
                \end{pmatrix},\
\mathbf{R_A} = \begin{pmatrix}
                0 & r \\ 
                0 & 0
                \end{pmatrix},\
\mathbf{R_D} = \begin{pmatrix}
                0 & 0 \\ 
                0 & r
                \end{pmatrix}.    
\end{equation}
Similarly we remove the repressor bound state from $\mathbf{R_A}
\langle\vec{m}^0\rangle$ and take the $k_R^+\rightarrow 0$ limit, which
simplifies to 
\begin{equation}
\mathbf{R_A} \langle\vec{m}^0\rangle
= r \frac{k_P^+ } {k_P^+ + k_P^- + r}
\begin{pmatrix} 1 \\ 0 \end{pmatrix}.
\end{equation}
Then we must compute $\langle\vec{m}\rangle$ from Eq.~\ref{eq:vecm_appdx_ref},
which with these matrices reads
\begin{equation}
(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} - \gamma\mathbf{I})
    \langle\vec{m}\rangle
= \begin{pmatrix}
    -k_P^+ -\gamma & k_P^- + r\\
    k_P^+          &-k_P^- - r - \gamma
    \end{pmatrix}
    \begin{pmatrix} m_E \\ m_P \end{pmatrix}
= r \frac{k_P^+ } {k_P^+ + k_P^- + r}
    \begin{pmatrix} 1 \\ 0 \end{pmatrix},
\end{equation}
where we labeled the components of $\langle\vec{m}\rangle$ as $m_E$ and $m_P$,
since they are the mean mRNA counts conditional upon the system residing in the
empty or polymerase bound states, respectively. Unlike for
$\langle\vec{m}^0\rangle$, the rows of this matrix are linearly independent so
we simply solve this matrix equation as is. We can immediately eliminate $m_E$
since $m_E = m_P (k_P^- + r + \gamma)/k_P^+$ from the second row, and
substituting into the first row gives an equation for $m_P$ alone, which is
\begin{equation}
\left[-(k_P^+ + \gamma)(k_P^- + r + \gamma) + k_P^+(k_P^- + r)\right] m_P
= - \frac{r (k_P^+)^2}{k_P^+ + k_P^- + r}.
\end{equation}
Expanding the products cancels several terms, and solving for $m_P$ gives
\begin{equation}
m_P = \frac{r (k_P^+)^2}
            {\gamma(k_P^+ + k_P^- + r)(k_P^+ + k_P^- + r + \gamma)}.
\end{equation}
Note then that $\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}\rangle = rm_P$.
We also need the constitutive limit of $\langle{m}\rangle$ from
Eq.~\ref{eq:model2_meanm_appdx}, again found by taking $k_R^+\rightarrow0$,
which is
\begin{equation}
\langle{m}\rangle = \frac{r}{\gamma} \frac{k_P^+ } {k_P^+ + k_P^- + r}
\end{equation}
and substituting this along with
$\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}\rangle = rm_P$ into
Eq.~\ref{eq:fano_appdx_ref} for the Fano factor $\nu$, we find
\begin{equation}
\nu = 1 - \frac{r}{\gamma} \frac{k_P^+ } {k_P^+ + k_P^- + r}
    + \frac{r}{\gamma}\frac{r (k_P^+)^2}{\gamma(k_P^+ + k_P^- + r)
                                        (k_P^+ + k_P^- + r + \gamma)}
\left(\frac{r}{\gamma} \frac{k_P^+ } {k_P^+ + k_P^- + r}\right)^{-1}.
\end{equation}
This simplifies to
\begin{equation}
\nu = 1 - \frac{r}{\gamma}
    \left(
        \frac{k_P^+ } {k_P^+ + k_P^- + r}
        - \frac{k_P^+ } {k_P^+ + k_P^- + r + \gamma}
    \right),
\end{equation}
which further simplifies to
\begin{equation}
\nu = 1 - \frac{r k_P^+ } {(k_P^+ + k_P^- + r)(k_P^+ + k_P^- + r + \gamma)},
\end{equation}
exactly Eq.~\ref{eq:model2_fano} in the main text.

@@

Mathematically this model is very similar to models 1 and 2. Like model 1, the
matrix $R$ describing transcription initiation is diagonal, namely
\begin{equation}
\mathbf{R} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & 0 \\ 
                0 & 0 & r
        \end{pmatrix}.
\end{equation}
The master equation takes verbatim the same form as it did for model 1,
Eq.~\ref{eq:2state_rep_cme}. Meanwhile the promoter transition matrix
$\mathbf{K}$ is the same as Eq.~\ref{eq:3state_cme_matrices_pt1} from model 2,
although we relabel the rate constants from $k_P^\pm$ to $k^\pm$ to reiterate
that these are not simply RNAP binding and unbinding rates.

>>>>

Having derived an example chemical master equation we now focus on writing a
general matrix form for the kinetic models 1-4 shown in
Figure~\ref{fig1:means_cartoons}(C) in the main text. In each of these four
models, the natural mathematicization of their cartoons is as a chemical master
equation. For model 1 we have the master equation
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma mp_U(m,t)}^{m \rightarrow m - 1}.
\label{eq:poisson_promoter_cme_appdx}
\end{split}
\end{align}
Here $p_R(m,t)$ and $p_U(m,t)$ are the probabilities of finding the system with
$m$ mRNA molecules at time $t$ either in the repressor bound or unbound states,
respectively. $r$ is the probability per unit time that a transcript will be
initiated when the repressor is unbound, and $\gamma$ is the probability per
unit time for a given mRNA to be degraded. $k_R^-$ is the probability per unit
time that a bound repressor will unbind, while $k_R^+$ is the probability per
unit time that an unbound operator will become bound by a repressor. Assuming
mass action kinetics, $k_R^+$ is proportional to repressor copy number $R$.

@@

Although we derived Eq.~\ref{eq:generic_cme_appdx} for the particular case of
kinetic model 2 in Figure~\ref{fig1:means_cartoons}, in fact the chemical master
equations for all of the kinetic models in Figure~\ref{fig1:means_cartoons}
except for model 5 can be cast in this form. (We treat model 5 separately in
Appendix~\ref{sec:gen_fcn_appdx}.) Model 3 introduces no new subtleties beyond
model 2 and Eq.~\ref{eq:generic_cme_appdx} applies equally well to it, simply
with different matrices of dimension four instead of three. Models 1 and 4 are
both Eq.~\ref{eq:generic_cme_appdx} except that $\mathbf{R_D} = \mathbf{R_A}
\equiv \mathbf{R}$, since in these two models transcription initiation events do
not change promoter state.

@@

\subsubsection{Fano factor}
To compute the Fano factor, we first remove the repressor bound state from the
matrices describing the model, which reduce to
\begin{equation}
\mathbf{K} = \begin{pmatrix}
                -k_P^+ & k_P^- \\
                 k_P^+ &-k_P^- 
                \end{pmatrix},\
\mathbf{R_A} = \begin{pmatrix}
                0 & r \\ 
                0 & 0
                \end{pmatrix},\
\mathbf{R_D} = \begin{pmatrix}
                0 & 0 \\ 
                0 & r
                \end{pmatrix}.    
\end{equation}
Similarly we remove the repressor bound state from $\mathbf{R_A}
\langle\vec{m}^0\rangle$ and take the $k_R^+\rightarrow 0$ limit, which
simplifies to 
\begin{equation}
\mathbf{R_A} \langle\vec{m}^0\rangle
= r \frac{k_P^+ } {k_P^+ + k_P^- + r}
\begin{pmatrix} 1 \\ 0 \end{pmatrix}.
\end{equation}
Then we must compute $\langle\vec{m}\rangle$ from Eq.~\ref{eq:vecm_appdx_ref},
which with these matrices reads
\begin{equation}
(\mathbf{K} - \mathbf{R_D} + \mathbf{R_A} - \gamma\mathbf{I})
    \langle\vec{m}\rangle
= \begin{pmatrix}
    -k_P^+ -\gamma & k_P^- + r\\
    k_P^+          &-k_P^- - r - \gamma
    \end{pmatrix}
    \begin{pmatrix} m_E \\ m_P \end{pmatrix}
= r \frac{k_P^+ } {k_P^+ + k_P^- + r}
    \begin{pmatrix} 1 \\ 0 \end{pmatrix},
\end{equation}
where we labeled the components of $\langle\vec{m}\rangle$ as $m_E$ and $m_P$,
since they are the mean mRNA counts conditional upon the system residing in the
empty or polymerase bound states, respectively. Unlike for
$\langle\vec{m}^0\rangle$, the rows of this matrix are linearly independent so
we simply solve this matrix equation as is. We can immediately eliminate $m_E$
since $m_E = m_P (k_P^- + r + \gamma)/k_P^+$ from the second row, and
substituting into the first row gives an equation for $m_P$ alone, which is
\begin{equation}
\left[-(k_P^+ + \gamma)(k_P^- + r + \gamma) + k_P^+(k_P^- + r)\right] m_P
= - \frac{r (k_P^+)^2}{k_P^+ + k_P^- + r}.
\end{equation}
Expanding the products cancels several terms, and solving for $m_P$ gives
\begin{equation}
m_P = \frac{r (k_P^+)^2}
            {\gamma(k_P^+ + k_P^- + r)(k_P^+ + k_P^- + r + \gamma)}.
\end{equation}
Note then that $\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}\rangle = rm_P$.
We also need the constitutive limit of $\langle{m}\rangle$ from
Eq.~\ref{eq:model2_meanm_appdx}, again found by taking $k_R^+\rightarrow0$,
which is
\begin{equation}
\langle{m}\rangle = \frac{r}{\gamma} \frac{k_P^+ } {k_P^+ + k_P^- + r}
\end{equation}
and substituting this along with
$\vec{1}^\dagger\cdot\mathbf{R_A}\langle\vec{m}\rangle = rm_P$ into
Eq.~\ref{eq:fano_appdx_ref} for the Fano factor $\nu$, we find
\begin{equation}
\nu = 1 - \frac{r}{\gamma} \frac{k_P^+ } {k_P^+ + k_P^- + r}
    + \frac{r}{\gamma}\frac{r (k_P^+)^2}{\gamma(k_P^+ + k_P^- + r)
                                        (k_P^+ + k_P^- + r + \gamma)}
\left(\frac{r}{\gamma} \frac{k_P^+ } {k_P^+ + k_P^- + r}\right)^{-1}.
\end{equation}
This simplifies to
\begin{equation}
\nu = 1 - \frac{r}{\gamma}
    \left(
        \frac{k_P^+ } {k_P^+ + k_P^- + r}
        - \frac{k_P^+ } {k_P^+ + k_P^- + r + \gamma}
    \right),
\end{equation}
which further simplifies to
\begin{equation}
\nu = 1 - \frac{r k_P^+ } {(k_P^+ + k_P^- + r)(k_P^+ + k_P^- + r + \gamma)}.
\end{equation}

@@

Mathematically this model is very similar to models 1 and 2. Like model 1, the
matrix $R$ describing transcription initiation is diagonal, namely
\begin{equation}
\mathbf{R} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & 0 \\ 
                0 & 0 & r
        \end{pmatrix}.
\end{equation}
The master equation takes verbatim the same form as it did for model 1.
Meanwhile the promoter transition matrix $\mathbf{K}$ is the same as
Eq.~\ref{eq:3state_cme_matrices_pt1} from model 2, although we relabel the rate
constants from $k_P^\pm$ to $k^\pm$ to reiterate that these are not simply RNAP
binding and unbinding rates.

### 2020-10-16 doc/resubmission/appendix_03_bayesian_inference.tex
fix ?? issues

As suggested by one of our reviewers, in order for readers to judge the
agreement between our predictions and the experimental data for the regulated
case, we include Figure~\ref{figS:eg_histograms} that plot the same data as
Figure~\ref{fig4:repressed_post_full}(C), but rather than showing ECDF, we show
histograms of the individual distributions.

>>>>

As suggested by one of our reviewers, in order for readers to judge the
agreement between our predictions and the experimental data for the regulated
case, we include Figure~\ref{figS:reg_histograms} that plot the same data as
Figure~\ref{fig4:repressed_post_full}(C), but rather than showing ECDF, we show
histograms of the individual distributions.
